Unshuffling Permutations

### 2016-02-10 submission.tex
Modif LATIN16

@@

\begin{abstract}
    A permutation is said to be a square if it can be obtained by
    shuffling two order-isomorphic patterns. The definition is intended
    to be the natural counterpart to the ordinary shuffle of words and
    languages. In this paper, we tackle the problem of recognizing square
    permutations from both the point of view of algebra and algorithms.
    On the one hand, we present some algebraic and combinatorial
    properties of the shuffle product of permutations. We follow an
    unusual line consisting in defining the shuffle of permutations by
    means of an unshuffling operator, known as a coproduct. This
    strategy allows to obtain easy proofs for algebraic and combinatorial
    properties of our shuffle product. We besides exhibit a bijection
    between square $(213,231)$-avoiding permutations and square binary
    words. On the other hand, using a containment-free perfect matching
    criterion, we prove that recognizing square permutations is \NPC.
\end{abstract}

@@

Let us now provide some definitions about graphs and matching that are used in
the sequel. For a graph $G$, we denote $\VS{G}$ as the set of vertices
and $\ES{G}$ as the set of edges of $G$. Let $u = u_1 u_2 \dots u_n$ be
a word. The \emph{clique graph associated with $u$}, denoted by $K_u$, is
the complete graph on $\VS{K_u} = \{u_1, u_2, \ldots, u_n\}$. The
structure of this underlying graph is linear, \emph{i.e.}, the set of
its vertices is equipped with a natural total order $<$ defined by
$u_i < u_j$ if and only if $i < j$. For a linear graph, we write
$(u_i, u_j)$ instead of $\{u_i, u_j\}$ for an edge if it is clear from
the context that $i < j$. Recall that two edges of a graph are
\emph{independent} if they do not share a common vertex, and that a
\emph{matching} $\mathcal{M}$ in $G$ is a set of pairwise independent
edges. A matching is \emph{perfect} if it covers all the vertices of the
graph.
For an edge $e = \{u,v\}$, we shall occasionally say that $u$ and $v$
are \emph{involved} in $e$.
In case the set of vertices is equipped with a total order, a
matching $\mathcal{M}$ is said to be \emph{containment-free} if there do
not exist (independent) edges $(u_i, u_j)$ and $(u_k, u_\ell)$ in
$\mathcal{M}$ such that $u_i < u_k < u_\ell < u_j$
%or $u_k < u_i < u_j < u_\ell$
(\emph{i.e.}, $\mathcal{M}$ avoids~\MWithContainment).

@@

This section is devoted to proving hardness of recognizing square
permutations. In the same way as happens with words, we shall use a
linear graph framework where deciding whether a permutation is a square
reduces to computing some specific matching in the associated linear
graph~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}. It reads as follows
(see Fig.\ref{fig:example containment-free matching}).
\begin{figure}[ht!]
  \centering
  \begin{tikzpicture}[scale=0.5,inner sep=2pt,node distance=1cm]
    \draw node (K) {$K_\pi$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=K]   (U01) {$2$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U01] (U02) {$A$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U02] (U03) {$4$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U03] (U04) {$7$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U04] (U05) {$1$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U05] (U06) {$9$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U06] (U07) {$5$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U07] (U08) {$3$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U08] (U09) {$8$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U09] (U10) {$6$};
    \draw [thick]
      (U01.north) .. controls ($ (U01.north) + (0,2.25) $) and ($ (U03.north) + (0,2.25) $) .. (U03.north);
    \draw [thick]
      (U02.north) .. controls ($ (U02.north) + (0,2.25) $) and ($ (U06.north) + (0,2.25) $) .. (U06.north);
    \draw [thick]
      (U04.north) .. controls ($ (U04.north) + (0,2.5) $) and ($ (U07.north) + (0,2.5) $) .. (U07.north);
    \draw [thick]
      (U05.north) .. controls ($ (U05.north) + (0,2.5) $) and ($ (U08.north) + (0,2.5) $) .. (U08.north);
    \draw [thick]
      (U09.north) .. controls ($ (U09.north) + (0,2.5) $) and ($ (U10.north) + (0,2.5) $) .. (U10.north);
  \end{tikzpicture}
  \caption{\label{fig:example containment-free matching}%
    The linear graph $K_\pi$ associated with the permutation
    $\pi = 2A47195386$ together with
    a containment-free perfect matching $\mathcal{M}$
    such that for any two distinct edges $(a, b)$ and
    $(a', b')$ in $\mathcal{M}$ we have $\pi(a) < \pi(a')$ if and only
    if $\pi(b) < \pi(b')$.
    From $\mathcal{M}$, it follows that $\pi$ is square as it is in
    the shuffle of $2A718$ and $49536$, both being
    order-isomorphic to $25314$.
  }% end caption
\end{figure}

>>>>

\newcommand{\Crossing}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']%,>=stealth',->]
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U3.north) + (0,4) $) .. (U3.north);
\draw [] (U2.north) .. controls ($ (U2.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\end{tikzpicture}
}% end \Crossing

\newcommand{\LabeledCrossingLL}[5]{%
\begin{tikzpicture}[scale=#1,inner sep=.8pt,node distance=.45cm,>=latex',
                    text height=1.5ex,text depth=.25ex]
\draw node [] (U1)               {#2};
\draw node [] [right of=U1] (U2) {#3};
\draw node [] [right of=U2] (U3) {#4};
\draw node [] [right of=U3] (U4) {#5};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,1.5) $) and ($ (U3.north) + (0,1.5) $) .. (U3.north);
\draw [->] (U2.north) .. controls ($ (U2.north) + (0,1.5) $) and ($ (U4.north) + (0,1.5) $) .. (U4.north);
\end{tikzpicture}
}% end \LabeledCrossingLL

\newcommand{\CrossingLL}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']%,>=stealth',->]
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,2) $) and ($ (U3.north) + (0,2) $) .. (U3.north);
\draw [->] (U2.north) .. controls ($ (U2.north) + (0,2) $) and ($ (U4.north) + (0,2) $) .. (U4.north);
\end{tikzpicture}
}% end \CrossingLL

\newcommand{\LabeledCrossingLR}[5]{%
\begin{tikzpicture}[scale=#1,inner sep=.8pt,node distance=.45cm,>=latex',
                    text height=1.5ex,text depth=.25ex]
\draw node [] (U1)               {#2};
\draw node [] [right of=U1] (U2) {#3};
\draw node [] [right of=U2] (U3) {#4};
\draw node [] [right of=U3] (U4) {#5};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,1.5) $) and ($ (U3.north) + (0,1.5) $) .. (U3.north);
\draw [<-] (U2.north) .. controls ($ (U2.north) + (0,1.5) $) and ($ (U4.north) + (0,1.5) $) .. (U4.north);
\end{tikzpicture}
}% end \LabeledCrossingLR

\newcommand{\CrossingLR}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']%,>=stealth',->]
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U3.north) + (0,4) $) .. (U3.north);
\draw [<-] (U2.north) .. controls ($ (U2.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\end{tikzpicture}
}% end \CrossingLR

\newcommand{\LabeledCrossingRL}[5]{%
\begin{tikzpicture}[scale=#1,inner sep=.8pt,node distance=.45cm,>=latex',
                    text height=1.5ex,text depth=.25ex]
\draw node [] (U1)               {#2};
\draw node [] [right of=U1] (U2) {#3};
\draw node [] [right of=U2] (U3) {#4};
\draw node [] [right of=U3] (U4) {#5};
\draw [<-] (U1.north) .. controls ($ (U1.north) + (0,1.5) $) and ($ (U3.north) + (0,1.5) $) .. (U3.north);
\draw [->] (U2.north) .. controls ($ (U2.north) + (0,1.5) $) and ($ (U4.north) + (0,1.5) $) .. (U4.north);
\end{tikzpicture}
}% end \LabeledCrossingRL

\newcommand{\CrossingRL}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']%,>=stealth',->]
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [<-] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U3.north) + (0,4) $) .. (U3.north);
\draw [->] (U2.north) .. controls ($ (U2.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\end{tikzpicture}
}% end \CrossingRL

\newcommand{\LabeledCrossingRR}[5]{%
\begin{tikzpicture}[scale=#1,inner sep=.8pt,node distance=.45cm,>=latex',
                    text height=1.5ex,text depth=.25ex]
\draw node [] (U1)               {#2};
\draw node [] [right of=U1] (U2) {#3};
\draw node [] [right of=U2] (U3) {#4};
\draw node [] [right of=U3] (U4) {#5};
\draw [<-] (U1.north) .. controls ($ (U1.north) + (0,1.5) $) and ($ (U3.north) + (0,1.5) $) .. (U3.north);
\draw [<-] (U2.north) .. controls ($ (U2.north) + (0,1.5) $) and ($ (U4.north) + (0,1.5) $) .. (U4.north);
\end{tikzpicture}
}% end \LabeledCrossingRR

\newcommand{\CrossingRR}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']%,>=stealth',->]
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U3.north) + (0,4) $) .. (U3.north);
\draw [->] (U2.north) .. controls ($ (U2.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\end{tikzpicture}
}% end \CrossingRR

\newcommand{\Inclusion}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\draw [] (U2.north) .. controls ($ (U2.north) + (0,3) $) and ($ (U3.north) + (0,3) $) .. (U3.north);
\end{tikzpicture}
}% end \Inclusion

\newcommand{\LabeledInclusionLL}[5]{%
\begin{tikzpicture}[scale=#1,inner sep=.8pt,node distance=.45cm,>=latex',
                    text height=1.5ex,text depth=.25ex]
\draw node [] (U1)               {#2};
\draw node [] [right of=U1] (U2) {#3};
\draw node [] [right of=U2] (U3) {#4};
\draw node [] [right of=U3] (U4) {#5};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,2) $) and ($ (U4.north) + (0,2) $) .. (U4.north);
\draw [->] (U2.north) .. controls ($ (U2.north) + (0,1) $) and ($ (U3.north) + (0,1) $) .. (U3.north);
\end{tikzpicture}
}% end \LabeledInclusionLL

\newcommand{\InclusionLL}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
+\draw [->] (U4.north) .. controls ($ (U4.north) + (0,4) $) and ($ (U1.north) + (0,4) $) .. (U1.north);
+\draw [->] (U3.north) .. controls ($ (U3.north) + (0,3) $) and ($ (U2.north) + (0,3) $) .. (U2.north);
\end{tikzpicture}
}% end \InclusionLL

\newcommand{\LabeledInclusionLR}[5]{%
\begin{tikzpicture}[scale=#1,inner sep=.8pt,node distance=.45cm,>=latex',
                    text height=1.5ex,text depth=.25ex]
\draw node [] (U1)               {#2};
\draw node [] [right of=U1] (U2) {#3};
\draw node [] [right of=U2] (U3) {#4};
\draw node [] [right of=U3] (U4) {#5};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,2) $) and ($ (U4.north) + (0,2) $) .. (U4.north);
\draw [<-] (U2.north) .. controls ($ (U2.north) + (0,1) $) and ($ (U3.north) + (0,1) $) .. (U3.north);
\end{tikzpicture}
}% end \LabeledInclusionLR

\newcommand{\InclusionLR}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\draw [<-] (U2.north) .. controls ($ (U2.north) + (0,3) $) and ($ (U3.north) + (0,3) $) .. (U3.north);
\end{tikzpicture}
}% end \InclusionLR

\newcommand{\LabeledInclusionRL}[5]{%
\begin{tikzpicture}[scale=#1,inner sep=.8pt,node distance=.45cm,>=latex',
                    text height=1.5ex,text depth=.25ex]
\draw node [] (U1)               {#2};
\draw node [] [right of=U1] (U2) {#3};
\draw node [] [right of=U2] (U3) {#4};
\draw node [] [right of=U3] (U4) {#5};
\draw [<-] (U1.north) .. controls ($ (U1.north) + (0,2) $) and ($ (U4.north) + (0,2) $) .. (U4.north);
\draw [->] (U2.north) .. controls ($ (U2.north) + (0,1) $) and ($ (U3.north) + (0,1) $) .. (U3.north);
\end{tikzpicture}
}% end \LabeledInclusionRL

\newcommand{\InclusionRL}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [<-] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\draw [->] (U2.north) .. controls ($ (U2.north) + (0,3) $) and ($ (U3.north) + (0,3) $) .. (U3.north);
\end{tikzpicture}
}% end \InclusionRL

\newcommand{\LabeledInclusionRR}[5]{%
\begin{tikzpicture}[scale=#1,inner sep=.8pt,node distance=.45cm,>=latex',
                    text height=1.5ex,text depth=.25ex]
\draw node [] (U1)               {#2};
\draw node [] [right of=U1] (U2) {#3};
\draw node [] [right of=U2] (U3) {#4};
\draw node [] [right of=U3] (U4) {#5};
\draw [<-] (U1.north) .. controls ($ (U1.north) + (0,2) $) and ($ (U4.north) + (0,2) $) .. (U4.north);
\draw [<-] (U2.north) .. controls ($ (U2.north) + (0,1) $) and ($ (U3.north) + (0,1) $) .. (U3.north);
\end{tikzpicture}
}% end \LabeledInclusionRR

\newcommand{\InclusionRR}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\draw [->] (U2.north) .. controls ($ (U2.north) + (0,3) $) and ($ (U3.north) + (0,3) $) .. (U3.north);
\end{tikzpicture}
}% end \InclusionRR

\newcommand{\Precedence}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U2.north) + (0,4) $) .. (U2.north);
\draw [] (U3.north) .. controls ($ (U3.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\end{tikzpicture}
}% end \Precedence

\newcommand{\LabeledPrecedenceLL}[5]{%
\begin{tikzpicture}[scale=#1,inner sep=.8pt,node distance=.45cm,>=latex',
                    text height=1.5ex,text depth=.25ex]
\draw node [] (U1)               {#2};
\draw node [] [right of=U1] (U2) {#3};
\draw node [] [right of=U2] (U3) {#4};
\draw node [] [right of=U3] (U4) {#5};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,1.5) $) and ($ (U2.north) + (0,1.5) $) .. (U2.north);
\draw [->] (U3.north) .. controls ($ (U3.north) + (0,1.5) $) and ($ (U4.north) + (0,1.5) $) .. (U4.north);
\end{tikzpicture}
}% end \PrecedenceLL

\newcommand{\PrecedenceLL}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U2.north) + (0,4) $) .. (U2.north);
\draw [->] (U3.north) .. controls ($ (U3.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\end{tikzpicture}
}% end \PrecedenceLL

\newcommand{\LabeledPrecedenceLR}[5]{%
\begin{tikzpicture}[scale=#1,inner sep=.8pt,node distance=.45cm,>=latex',
                    text height=1.5ex,text depth=.25ex]
\draw node [] (U1)               {#2};
\draw node [] [right of=U1] (U2) {#3};
\draw node [] [right of=U2] (U3) {#4};
\draw node [] [right of=U3] (U4) {#5};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,1.5) $) and ($ (U2.north) + (0,1.5) $) .. (U2.north);
\draw [<-] (U3.north) .. controls ($ (U3.north) + (0,1.5) $) and ($ (U4.north) + (0,1.5) $) .. (U4.north);
\end{tikzpicture}
}% end \PrecedenceLR

\newcommand{\PrecedenceLR}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U2.north) + (0,4) $) .. (U2.north);
\draw [<-] (U3.north) .. controls ($ (U3.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\end{tikzpicture}
}% end \PrecedenceLR

\newcommand{\LabeledPrecedenceRL}[5]{%
\begin{tikzpicture}[scale=#1,inner sep=.8pt,node distance=.45cm,>=latex',
                    text height=1.5ex,text depth=.25ex]
\draw node [] (U1)               {#2};
\draw node [] [right of=U1] (U2) {#3};
\draw node [] [right of=U2] (U3) {#4};
\draw node [] [right of=U3] (U4) {#5};
\draw [<-] (U1.north) .. controls ($ (U1.north) + (0,1.5) $) and ($ (U2.north) + (0,1.5) $) .. (U2.north);
\draw [->] (U3.north) .. controls ($ (U3.north) + (0,1.5) $) and ($ (U4.north) + (0,1.5) $) .. (U4.north);
\end{tikzpicture}
}% end \PrecedenceRL

\newcommand{\PrecedenceRL}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [<-] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U2.north) + (0,4) $) .. (U2.north);
\draw [->] (U3.north) .. controls ($ (U3.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\end{tikzpicture}
}% end \PrecedenceRL

\newcommand{\LabeledPrecedenceRR}[5]{%
\begin{tikzpicture}[scale=#1,inner sep=.8pt,node distance=.45cm,>=latex',
                    text height=1.5ex,text depth=.25ex]
\draw node [] (U1)               {#2};
\draw node [] [right of=U1] (U2) {#3};
\draw node [] [right of=U2] (U3) {#4};
\draw node [] [right of=U3] (U4) {#5};
\draw [<-] (U1.north) .. controls ($ (U1.north) + (0,1.5) $) and ($ (U2.north) + (0,1.5) $) .. (U2.north);
\draw [<-] (U3.north) .. controls ($ (U3.north) + (0,1.5) $) and ($ (U4.north) + (0,1.5) $) .. (U4.north);
\end{tikzpicture}
}% end \PrecedenceRR

\newcommand{\PrecedenceRR}{%
\begin{tikzpicture}[yscale=0.08,xscale=0.09,inner sep=.8pt,node distance=.25cm,>=latex']
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw [<-] (U1.north) .. controls ($ (U1.north) + (0,4) $) and ($ (U2.north) + (0,4) $) .. (U2.north);
\draw [<-] (U3.north) .. controls ($ (U3.north) + (0,4) $) and ($ (U4.north) + (0,4) $) .. (U4.north);
\end{tikzpicture}
}% end \PrecedenceRR

@@

\begin{abstract}
    A permutation is said to be a square if it can be obtained by
    shuffling two order-isomorphic patterns. The definition is intended
    to be the natural counterpart to the ordinary shuffle of words and
    languages. In this paper, we tackle the problem of recognizing square
    permutations from both the point of view of algebra and algorithms.
    On the one hand, we present some algebraic and combinatorial
    properties of the shuffle product of permutations. We follow an
    unusual line consisting in defining the shuffle of permutations by
    means of an unshuffling operator, known as a coproduct. This
    strategy allows to obtain easy proofs for algebraic and combinatorial
    properties of our shuffle product. We besides exhibit a bijection
    between square $(213,231)$-avoiding permutations and square binary
    words. On the other hand, by using a pattern avoidance criterion on
    oriented perfect matchings, we prove that recognizing square 
    permutations is \NPC.
\end{abstract}

@@

Let us now provide some definitions about graphs and oriented perfect 
matchings that are used in the sequel. If $G$ is an oriented graph 
without loops, two different edges of $G$ are {\em independent} if they 
do not share any common vertex. We say that $G$ is an 
{\em oriented matching} if all edges of $G$ are pairwise independent. 
Moreover, $G$ is {\em perfect} if all vertices of $G$ belong to at least 
one arc. For any word $u = u_1 u_2 \dots u_n$, an 
{\em oriented perfect matching on $u$} is an oriented perfect matching 
$\mathcal{M}$ on the set of vertices $\{u_1, u_2, \dots, u_n\}$. In the 
sequel, we shall consider a natural notion of pattern avoidance in 
oriented perfect matchings on words. For instance, an oriented perfect
matching $\mathcal{M}$ of a word $u$ {\em admits an occurrence} of the 
pattern $\CrossingRL$ if there are four positions $i < j < k < \ell$ in 
$u$ such that $(u_k, u_i)$ and $(u_j, u_\ell)$ are arcs of $\mathcal{M}$. 
When $\mathcal{M}$ does not admit any occurrence of a pattern 
$\mathcal{P}$, we say that $\mathcal{M}$ {\em avoids}~$\mathcal{P}$.

% Let us now provide some definitions about graphs and oriented matching 
% that are used in the sequel. For a graph $G$, we denote $\VS{G}$ as the 
% set of vertices and $\ES{G}$ as the set of edges of $G$. Let 
% $u = u_1 u_2 \dots u_n$ be a word. The \emph{clique graph associated with $u$},
% denoted by $K_u$, is the complete graph on 
% $\VS{K_u} = \{u_1, u_2, \ldots, u_n\}$. The structure of this underlying 
% graph is linear, \emph{i.e.}, the set of
% its vertices is equipped with a natural total order $<$ defined by
% $u_i < u_j$ if and only if $i < j$. For a linear graph, we write
% $(u_i, u_j)$ instead of $\{u_i, u_j\}$ for an edge if it is clear from
% the context that $i < j$. Recall that two edges of a graph are
% \emph{independent} if they do not share a common vertex, and that a
% \emph{matching} $\mathcal{M}$ in $G$ is a set of pairwise independent
% edges. A matching is \emph{perfect} if it covers all the vertices of the
% graph.
% For an edge $e = \{u,v\}$, we shall occasionally say that $u$ and $v$
% are \emph{involved} in $e$.
% In case the set of vertices is equipped with a total order, a
% matching $\mathcal{M}$ is said to be \emph{containment-free} if there do
% not exist (independent) edges $(u_i, u_j)$ and $(u_k, u_\ell)$ in
% $\mathcal{M}$ such that $u_i < u_k < u_\ell < u_j$
% %or $u_k < u_i < u_j < u_\ell$
% (\emph{i.e.}, $\mathcal{M}$ avoids~\MWithContainment).

@@

This section is devoted to proving hardness of recognizing square
permutations. In the same way as happens with words, we shall use a
linear graph framework where deciding whether a permutation is a square
reduces to computing some specific matching in the associated linear
graph~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}. We have, however, 
to deal with oriented perfect matchings. The needed properties read 
as follows (see Fig.~\ref{fig:example containment-free matching}).
\begin{figure}[ht!]
    \centering
    \begin{tikzpicture}[scale=0.5,inner sep=2pt,node distance=1cm]
        \draw node [draw,circle,minimum size=.55cm,fill=black!10](U01){$1$};
        \draw node [draw,circle,minimum size=.55cm,fill=black!10,right of=U01]          
            (U02) {$8$};
        \draw node [draw,circle,minimum size=.55cm,right of=U02](U03){$3$};
        \draw node [draw,circle,minimum size=.55cm,fill=black!10,right of=U03] 
            (U04) {$9$};
        \draw node [draw,circle,minimum size=.55cm,fill=black!10,right of=U04]
            (U05) {$2$};
        \draw node [draw,circle,minimum size=.55cm,right of=U05](U06){$7$};
        \draw node [draw,circle,minimum size=.55cm,right of=U06](U07){$B$};
        \draw node [draw,circle,minimum size=.55cm,right of=U07](U08){$5$};
        \draw node [draw,circle,minimum size=.55cm,right of=U08](U09){$C$};
        \draw node [draw,circle,minimum size=.55cm,right of=U09](U10){$6$};
        \draw node [draw,circle,minimum size=.55cm,fill=black!10,right of=U10]
            (U11) {$A$};
        \draw node [draw,circle,minimum size=.55cm,fill=black!10,right of=U11]
            (U12) {$4$};
        \draw [thick,->]
        (U01.north) .. controls ($ (U01.north) + (0,2.25) $) 
            and ($ (U03.north) + (0,2.25) $) .. (U03.north);
        \draw [thick,->]
        (U02.north) .. controls ($ (U02.north) + (0,2.25) $) 
            and ($ (U06.north) + (0,2.25) $) .. (U06.north);
        \draw [thick,->]
        (U04.north) .. controls ($ (U04.north) + (0,2.5) $) 
            and ($ (U07.north) + (0,2.5) $) .. (U07.north);
        \draw [thick,->]
        (U05.north) .. controls ($ (U05.north) + (0,2.5) $) 
            and ($ (U08.north) + (0,2.5) $) .. (U08.north);
        \draw [thick,->]
        (U11.north) .. controls ($ (U11.north) + (0,2.5) $) 
            and ($ (U09.north) + (0,2.5) $) .. (U09.north);
        \draw [thick,->]
        (U12.north) .. controls ($ (U12.north) + (0,2.5) $) 
            and ($ (U10.north) + (0,2.5) $) .. (U10.north);
    \end{tikzpicture}
    \caption{\label{fig:example containment-free matching}%
        An oriented perfect matching $\mathcal{M}$ on the permutation
        $\pi = 183927B5C6A4$ satisfying the properties $\mathbf{P_1}$ 
        and $\mathbf{P_2}$. From $\mathcal{M}$, it follows that $\pi$ is 
        a square as it appears in the shuffle of $1892A4$ and $37B5C6$, 
        both being order-isomorphic to $145263$.
    }% end caption
\end{figure}

\begin{definition}[Property $\mathbf{P_1}$]
  \label{definition:Property P_1}
  Let $\pi$ be a permutation. An oriented perfect matching $\mathcal{M}$
  on $\pi$ is said to have property $\mathbf{P_1}$ if it avoids the 
  patterns \InclusionLL, \InclusionLR, \InclusionRL, 
  \InclusionRR, \CrossingLR, and \CrossingRL.
\end{definition}

\begin{definition}[Property $\mathbf{P_2}$]
  \label{definition:Property P_2}
  Let $\pi$ be a permutation. An oriented perfect matching $\mathcal{M}$
  on $\pi$ is said to have property $\mathbf{P_2}$ if, for any two 
  distinct arcs $(a, a')$ and $(b, b')$ in $\mathcal{M}$,
  we have $\pi(a) < \pi(b)$ if and only if $\pi(a') < \pi(b')$.
\end{definition}

The rationale for introducing properties $\mathbf{P_1}$ and $\mathbf{P_2}$
stems from the following lemma.

\begin{lemma}
  \label{lemma:matching}
  Let $\pi$ be a permutation. The following statements are equivalent:
  \begin{enumerate}
    \item The permutation $\pi$ is a square.
    \item There exists an oriented perfect matching $\mathcal{M}$
    on $\pi$ satisfying~$\mathbf{P_1}$ and~$\mathbf{P_2}$.
  \end{enumerate}
\end{lemma}

Let $\pi$ be a permutation. For the sake of clarity, we will say that a
bunch of consecutive positions $P$ of $\pi$ is \emph{above} (resp.
\emph{below}) above another bunch of consecutive positions $P'$ in $\pi$
if $\pi(i) > \pi(j)$ (resp. $\pi(i) < \pi(j)$) for every $i \in P$ and
every $j \in P'$. For example, $\sigma_1$ is above $\sigma_2$ (in an
equivalent manner, $\sigma_2$ is below $\sigma_1$) in
Fig.~\ref{subfig:increasing before and above decreasing}, whereas
$\sigma_1$ is below $\sigma_2$ (in an equivalent manner, $\sigma_2$ is
above $\sigma_1$) in Fig.~\ref{subfig:decreasing before and below
increasing}.
  \subfigure[%
    An increasing pattern before and above a decreasing pattern.
  ]{%
    \begin{tikzpicture}
      [
        scale=0.2,
        label/.style={anchor=base}
      % increasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,5.8) grid (5.2,11.2);
      \foreach \x/\y in {0/6,1/7,2/8,3/9,4/10,5/11} {
          \draw [fill=black] (\x,\y) circle (0.1);
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,-0.2) grid (11.2,5.2);
      \foreach \x/\y in {6/5,7/4,8/3,9/2,10/1,11/0} {
